---
title: "ONS Business Demography dataset 2017-22: what does it say about South Yorkshire?"
author: "Dan Olner"
format: 
  html:
    output-file: "ONS_business_demography"
execute:
  echo: false
  message: false
  warning: false
---


```{r, setup}
#Original code to work all this out is in same repo, "ONS_business_demography.R"

#https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  comment = '', fig.width = 9, fig.height = 5, dpi = 150
  # comment = '', fig.width = 13, fig.height = 8, dpi = 150
)

library(tidyverse)
library(sf)
library(tmap)
library(zoo)
source('../functions.R')

#List of separate dataframes for each key sheet from the original data
d <- readRDS('../data/businessdemog.rds')

#Combine all
bd <- d$births %>% 
  rename(count_births = count) %>% 
  left_join(
    d$deaths %>% select(region,year,count_deaths = count), by = c('region','year')
  ) %>% 
  left_join(
    d$active %>% select(region,year,count_active = count), by = c('region','year')
  ) %>%
  left_join(
    d$active10plus %>% select(region,year,count_active10plus = count), by = c('region','year')
  ) %>%
  left_join(
    d$highgrowth %>% select(region,year,count_highgrowth = count), by = c('region','year')
  ) %>%
  relocate(count_deaths, .after = count_births) %>% 
  relocate(count_active, .after = count_deaths) %>% 
  relocate(count_active10plus, .after = count_active) %>% 
  relocate(count_highgrowth, .after = count_active10plus) 

la22boundaries <- st_read("~/Dropbox/MapPolygons/UK/2022/Local_Authority_Districts_December_2022_UK", quiet = T) %>% mutate(LAD22NM = gsub(',','',LAD22NM)) %>% st_simplify(preserveTopology = T, dTolerance = 100)

#Made here: 
sy.geo <- readRDS('../localdata/sygeo.rds')


```



## Headlines

* **High growth firms in South Yorkshire and nationally (10+ employees and 20%+ employee growth av per year over 3 years):**
    + Since 2017, the **number and proportion of high growth firms has *dropped* almost everywhere** - but the **position of different regions _relative to each other_ have hardly changed.**
    + South Yorkshire (and Sheffield, Barnsley, Doncaster + Rotherham separately) have **remained at or near the bottom for proportion of high growth firms**, with Rotherham in particular dropping over time. (See @fig-highgrowth-raw for raw MCA numbers, @fig-highgrowth-percentoffirms10plus as percent of active & @fig-highgrowth-percentoffirms10plus-cores for the 4 places - **hover for preview** or see below.)

* **Business births and deaths:**
    + Nationally, business births have outnumbered deaths in every year in this data (though with exceptions, see map below) - but **firm deaths have steadily increased in last 4 or so years.**
    + **South Yorkshire is near the top of MCAs for births as a percent of active firms** - increasing over time, putting it in a higher group with West Midlands, Liverpool and Greater Manchester. SY has followed the national trend of increasing firm deaths as a percent of active firms, sitting in the middle rankings. (See @fig-birthsdeaths2axis_MCA_facet.)
    + @fig-birthsdeaths2axis_core_facet examines the same for **'core cities' in England/Scotland plus Barnsley, Doncaster & Rotherham (BDR)**. The stand-out here is **Doncaster** with high birth and death proportions compared both to other core cities and BDR. Since 2018-20, all four places **except Sheffield** have seen births *and* deaths relative to active firms *rise*. Sheffield is relatively **low for births** and, in this data, it has dropped to the **lowest birth value for the 2020-22 average.** (Note: only compared to the local authorities in this short list.)

## Maps

* **Four maps below give an overview of the national picture**:
    + @fig-birthdeathmap shows the **linear trend over time for firm births and deaths** nationally, expressed as % change per year (approx) with births in the top map and deaths below it (South Yorkshire overlaid in thicker black). Regional clustering is evident but - **apart from Doncaster's dropping death rate (which has been climbing again in more recent years)** - South Yorkshire doesn't stand out.
    + @fig-highgrowthchange is a map of the linear trend for **high growth firms** - again, showing approx. percent change per year. This map confirms the story that high growth firm numbers have been consistently dropping (red) but shows that **a patchwork of places in blue have seen high growth firm increases, with no obvious pattern**. South Yorkshire fits the pattern of high growth firm decline.
    + @fig-turnovermap shows firm 'turnover' (2020-22 average). **Turnover is a measure of churn**: births _plus_ deaths, divided by active firms. **Bands of higher turnover/churn** are present in the North, Midlands, across London and in South Wales, with **Doncaster standing out in the North** and the rest of South Yorkshire not especially low churn.
    + @fig-firmefficiencymap shows 'firm efficiency' - a **standardised measure of the balance of births & deaths**, where values of +100 would be 'all births no deaths', -100 would be 'all deaths no births', and zero would be 'births and deaths balance / cancel out'. This map compares two time points, showing how in the latter time period (bottom map, 2020-22) there has been a **marked shift to more deaths than births in the South**.


## The data used

* The [ONS Business Demography dataset](https://www.ons.gov.uk/businessindustryandtrade/business/activitysizeandlocation/datasets/businessdemographyreferencetable) contains counts of active firms, births, deaths, 'high growth' firms (20%+ employee growth over 3 years) and active firms with 10+ employees, from 2017 to 2022.
* Here, it has been harmonised to backcast 2022 local authority borders to create a time series not present in the original data.
* At local authority or MCA level, it does **not** break these counts down by sector (those are only at national level).
* R Code is [here](https://github.com/DanOlner/FirmAnalysis/blob/master/ONS_business_demography.R), quarto code for this report is [here](https://github.com/DanOlner/FirmAnalysis/blob/master/docs/QUARTO_ONS_businessdemography.qmd).




## The detail / plots / maps

### High growth firms

Raw high growth firm numbers have been declining almost everywhere in the UK since 2017. @fig-highgrowth-raw shows raw count of high growth firms for MCAs (thicker line is South Yorkshire).

However, most places haven't changed **relative** position very much over time, with one or two exceptions. 

```{r, fig-highgrowth-raw}
#| label: fig-highgrowth-raw
#| fig-cap: "For MCAs: Count of high growth firms 2017 to 2021. Log scale for size comparisons."

mcas.hg <- d$highgrowth %>%
  filter(!is.na(CAUTH24NM)) %>% 
  group_by(CAUTH24NM,year) %>%
  summarise(count = sum(count)) %>% 
  rename(MCA = CAUTH24NM)


ggplot(mcas.hg %>% mutate(SY = MCA == "South Yorkshire"), 
       aes(x = year, y = count, colour = fct_reorder(MCA,-count), size = SY)) +
  geom_line() +
  geom_point(size = 2) +
  scale_color_brewer(palette = 'Paired') +
  scale_y_log10() +
  scale_size_manual(values = c(0.5,2.5)) +
  labs(colour = 'MCA') +
  guides(size = F) 
  # ggtitle('Count of high growth firms in MCAs 2017 to 2021') +
  # theme(plot.title = element_text(hjust = 0.5, face = 'bold'))

```

The same overall pattern is visible in @fig-highgrowth-percentoffirms10plus, looking at high growth firms as a percent of active 10+ employee firms ("high growth firms" in ONS is also 10+ employees only). 

Thicker line is South Yorkshire. 

**Liverpool City Region** appears to be the only MCA bucking the trend.

Separate analysis shows it is mostly the drop in high growth firm count - not a large increase in active firms overall - causing this change. But note that firms can drop in/out of the "high growth" category with e.g. just one year of below-20% employee growth.


```{r, fig-highgrowth-percentoffirms10plus}
#| label: fig-highgrowth-percentoffirms10plus
#| fig-cap: "For MCAs: high growth firms as a percent of all active firms 10+ employees in MCAs, smoothed 3 yr average 2017/19 to 2020/22."

#Merge as new column to make division easy
hg_propof10plusfirms <- d$highgrowth %>% 
  rename(count_highgrowth = count) %>% 
  left_join(
    d$active10plus %>% select(region,year,count_active10plus = count), by = c('region','year')
  ) 

#Repeat plots
#Have to get percent here so it's after summing counts
mcas.hg.prop <- hg_propof10plusfirms %>%
  filter(!is.na(CAUTH24NM)) %>% 
  group_by(CAUTH24NM,year) %>%
  summarise(
    count_highgrowth = sum(count_highgrowth),
    count_active10plus = sum(count_active10plus)
    ) %>% 
  rename(MCA = CAUTH24NM) %>% 
  mutate(
    highgrowthfirms_aspercentof_firms10plusemployees = (count_highgrowth/count_active10plus) * 100,
    highgrowthfirms_aspercentof_firms10plusemployees_movingav = rollapply(highgrowthfirms_aspercentof_firms10plusemployees, 3, mean, align = 'center', fill = NA)
    )

#3 yr  
ggplot(mcas.hg.prop %>% mutate(SY = MCA == "South Yorkshire"), 
       aes(size = SY, x = year, y = highgrowthfirms_aspercentof_firms10plusemployees_movingav, colour = fct_reorder(MCA,-highgrowthfirms_aspercentof_firms10plusemployees_movingav))) +
  geom_line() +
  geom_point(size = 2) +
  scale_color_brewer(palette = 'Paired') +
  coord_cartesian(xlim = c(2018,2021)) + 
  scale_size_manual(values = c(0.5,2.5)) +
  labs(colour = 'MCA') +
  guides(size = F) +
  ylab('High growth firms as % of active 10+ employee firms')

```


@fig-highgrowth-percentoffirms10plus-cores also shows high growth firms as a percent of active firms but showing **Sheffield, Barnsley, Doncaster & Rotherham** separately (thicker lines) against English/Scottish **core cities** (minus Cardiff & Belfast).

The same story of stable positions against secular trend dropping is here, with S+BDR always near the bottom - but Rotherham more recently dropping more than others (and again, Liverpool the only place here goin the other way).

Using 16+ in employment (not shown here) doesn't change the relative position / direction of S+BDR.

Separate analysis again shows, for S+BDR, drops in raw 'high growth' firms against (smaller) rises in 10+ employee active firms.


```{r, fig-highgrowth-percentoffirms10plus-cores}
#| label: fig-highgrowth-percentoffirms10plus-cores
#| fig-cap: "For core cities plus Barnsley/Doncaster/Rotherham: High growth firms as a percent of all active firms 10+ employees, smoothed 3 yr average 2017/19 to 2020/22."

#Let's do same for core cities... each is already separate LA, don't need to summarise
#THIS LOOKS TOO VOLATILE, PROB NOT ENOUGH DATA FOR NON SMOOTHED
#Use moving av again
core.hg.prop <- hg_propof10plusfirms %>%
  filter(corecity == 'Core city' | grepl('barnsley|doncast|rotherh', region, ignore.case = T)) %>% 
  filter(!grepl('cardiff|Belfast', region, ignore.case = T)) %>% 
  mutate(highgrowthfirms_aspercentof_firms10plusemployees = (count_highgrowth/count_active10plus) * 100 ) %>% 
  group_by(region) %>% 
  mutate(
    count_highgrowth_movingav = rollapply(count_highgrowth, 3, mean, align = 'center', fill = NA),
    count_active10plus_movingav = rollapply(count_active10plus, 3, mean, align = 'center', fill = NA),
    highgrowthfirms_aspercentof_firms10plusemployees_movingav = rollapply(highgrowthfirms_aspercentof_firms10plusemployees, 3, mean, align = 'center', fill = NA)
  ) %>% 
  ungroup()
  
#Volatile, use moving av
ggplot(core.hg.prop%>% mutate(SBDR = grepl('barnsley|doncast|rotherh|sheffield',region,ignore.case=T)), 
       aes(size = SBDR, x = year, y = highgrowthfirms_aspercentof_firms10plusemployees_movingav, colour = fct_reorder(region,-highgrowthfirms_aspercentof_firms10plusemployees_movingav))) +
  geom_line() +
  geom_point(size = 2) +
  scale_color_brewer(palette = 'Paired') +
  coord_cartesian(xlim = c(2018,2021)) +
  labs(colour = 'Core cities (plus BDR)') +
  scale_size_manual(values = c(0.5,2.5)) +
  # scale_y_log10() +
  guides(size = F) +
  ylab('High growth firms as % of active 10+ employee firms')


```


### Births and deaths

Nationally, business births have outnumbered deaths in every year in this data - but **firm deaths have steadily increased in the last 4 or so years.**

@fig-birthsdeaths2axis_MCA_facet shows South Yorkshire near the top of MCAs for births as a percent of active firms - increasing over time, putting it in a higher group with West Midlands, Liverpool and Greater Manchester. SY has followed the national trend of increasing firm deaths as a percent of active firms, sitting in the middle rankings. 

```{r}
#| label: fig-birthsdeaths2axis_MCA_facet
#| fig-cap: "For MCAs: births vs deaths as a percent of all active firms, 3 year moving average 2017/19 to 2020/22."
#| fig-height: 7

bd.mca <- bd %>%
  filter(!is.na(CAUTH24NM)) %>% 
  group_by(CAUTH24NM,year) %>%
  summarise(
    count_births = sum(count_births),
    count_deaths = sum(count_deaths),
    count_active = sum(count_active)
    ) %>% 
  rename(MCA = CAUTH24NM) %>% 
  mutate(
    firmefficency = (count_births - count_deaths)/(count_births + count_deaths) * 100,#diffs here quite small so scale to 100
    turnover = (count_births + count_deaths)/(count_active),
    births_over_active_percent = ((count_births)/(count_active))*100,
    deaths_over_active_percent = ((count_deaths)/(count_active))*100
  ) %>% 
  group_by(MCA) %>% 
  mutate(
    firmefficency_movingav = rollapply(firmefficency, 3, mean, align = 'center', fill = NA),
    turnover_movingav = rollapply(turnover, 3, mean, align = 'center', fill = NA),
    births_over_active_percent_movingav = rollapply(births_over_active_percent, 3, mean, align = 'center', fill = NA),
    deaths_over_active_percent_movingav = rollapply(deaths_over_active_percent, 3, mean, align = 'center', fill = NA)
  ) %>% 
  ungroup()


#Let's do a faceted version of those last two with same scale, side by side if poss
bd.mca.birthdeath <- bd.mca %>% 
  ungroup() %>% 
  select(MCA,year,`Births as percent of active firms` = births_over_active_percent_movingav,
         `Deaths as percent of active firms` = deaths_over_active_percent_movingav,turnover_movingav) %>% 
  pivot_longer(cols = contains("percent of active firms"), names_to = "birthsdeaths", values_to = "percent") %>% 
  mutate(MCA = stringr::str_sub(MCA, start = 1, end=9))

ggplot(bd.mca.birthdeath %>% mutate(SY = grepl("South Y",MCA,ignore.case=T)), 
       aes(size = SY, x = year, y = percent, colour = fct_reorder(MCA,-turnover_movingav))) +
  geom_line() +
  geom_point(size = 2) +
  scale_color_brewer(palette = 'Paired') +
  scale_size_manual(values = c(0.5,2)) +
  coord_cartesian(xlim = c(2018,2021)) +
  facet_wrap(~birthsdeaths, nrow = 1) +
  labs(colour = 'MCA') +
  xlab('3 year moving average, mid-year-point') +
  guides(size = F)


```

@fig-birthsdeaths2axis_core_facet examines the same for **core cities plus BDR**. The stand-out here is **Doncaster** with high birth and death proportions compared both to other core cities and the other three SY places. Since 2018-20, all four places **except Sheffield** have seen births *and* deaths relative to active firms *rise*. Sheffield is relatively **low for births** and, in this data, it has dropped to the **lowest firm death value.** (Note: only compared to the local authorities in this short list.)


```{r}
#| label: fig-birthsdeaths2axis_core_facet
#| fig-cap: "For core cities plus BDR: births vs deaths as a percent of all active firms, 3 year moving average 2017/19 to 2020/22."
#| fig-height: 7


bd.core <- bd %>%
  # filter(corecity == 'Core city') %>% 
  filter(corecity == 'Core city' | grepl('barns|doncaster|rotherh',region,ignore.case=T)) %>% #version that includes other places in SY
  filter(!grepl('cardiff|belfast',region,ignore.case=T)) %>% 
  mutate(
    firmefficency = (count_births - count_deaths)/(count_births + count_deaths) * 100,#diffs here quite small so scale to 100
    turnover = (count_births + count_deaths)/(count_active),
    births_over_active_percent = ((count_births)/(count_active))*100,
    deaths_over_active_percent = ((count_deaths)/(count_active))*100
  ) %>% 
  group_by(region) %>% 
  mutate(
    #Smoothed variables
    highgrowth_movingav = rollapply(count_highgrowth, 3, mean, align = 'center', fill = NA),
    #derived measures
    firmefficency_movingav = rollapply(firmefficency, 3, mean, align = 'center', fill = NA),
    turnover_movingav = rollapply(turnover, 3, mean, align = 'center', fill = NA),
    births_over_active_percent_movingav = rollapply(births_over_active_percent, 3, mean, align = 'center', fill = NA),
    deaths_over_active_percent_movingav = rollapply(deaths_over_active_percent, 3, mean, align = 'center', fill = NA)
  ) %>% 
  ungroup()

#For facetting, make births and deaths long
bd.core.birthdeath <- bd.core %>% 
  ungroup() %>% 
  select(region,year,`Births as percent of active firms` = births_over_active_percent_movingav,
         `Deaths as percent of active firms` = deaths_over_active_percent_movingav,turnover_movingav) %>% 
  pivot_longer(cols = contains("percent of active firms"), names_to = "birthsdeaths", values_to = "percent") %>% 
  mutate(region = stringr::str_sub(region, start = 1, end=6))


ggplot(bd.core.birthdeath %>% mutate(SBDR = grepl('barnsl|doncas|roth|sheffi',region,ignore.case=T)), 
     aes(size = SBDR, x = year, y = percent, colour = fct_reorder(region,-turnover_movingav))) +
  geom_line() +
  geom_point(size = 2) +
  scale_color_brewer(palette = 'Paired') +
  scale_size_manual(values = c(0.5,2)) +
  coord_cartesian(xlim = c(2018,2021)) +
  facet_wrap(~birthsdeaths, nrow = 1) +
  labs(colour = 'MCA') +
  xlab('3 year moving average, mid-year-point') +
  guides(size = F)


```



## Maps

@fig-birthdeathmap shows the linear trends for business births (top map) and deaths (bottom) for the whole of the UK, expressed as approximate percentage change per year on average. South Yorkshire is marked with a thicker black line. 

The overall pattern of births is mixed and clustered across the UK - a swathe of the South dropping more than many other places. Increases and drops in deaths also cluster.

```{r, fig-birthdeathmap}
#| label: fig-birthdeathmap
#| fig-cap: "Linear trend in births (top map) and deaths (bottom map) of firms 2017-2022, expressed in approx. percent change per year"
#| fig-height: 17

birthslopes <- get_slope_and_se_safely(
    data = bd, 
    region,#slopes will be found within whatever grouping vars are added here
    y = log(count_births), x = year
    )
  
deathslopes <- get_slope_and_se_safely(
    data = bd, 
    region,#slopes will be found within whatever grouping vars are added here
    y = log(count_deaths), x = year
    )
  
bd.geo <- la22boundaries %>% 
  rename(region = LAD22NM) %>% 
  left_join(
    bd,
    by = c('region')
  ) %>% 
  left_join(
    birthslopes %>% rename(`Births % change / yr (approx)` = slope, birth_se = se),
    by = 'region'
  ) %>% 
  left_join(
    deathslopes %>% rename(`Deaths % change / yr (approx)` = slope, death_se = se),
    by = 'region'
  )

#Make slopes long so can facet
bd.geo.long <- bd.geo %>% 
  select(region,`Births % change / yr (approx)`,`Deaths % change / yr (approx)`) %>% 
  pivot_longer(
    cols = contains("approx"), names_to = "birthsdeathslopes", values_to = "approx_percent"
  ) %>% 
  mutate(approx_percent = approx_percent * 100)

tm_shape(bd.geo.long) +
  tm_polygons('approx_percent', n = 13, palette = "RdBu", border.alpha = 0.4, 
              title = "Firm births\n& deaths\nApprox %\nchange/yr") +
  tm_layout(title = '', legend.outside = T) +
  tm_facets(by = 'birthsdeathslopes', ncol = 1) +
  tm_shape(sy.geo) + #after facets or doesn't work
  tm_borders(lwd = 2, col = 'black') 




```


@fig-highgrowthchange is a map of the linear trend for **high growth firms** - again, showing approx. percent change per year. This map confirms the story that high growth firm numbers have been consistently dropping (red) but shows that **a patchwork of places in blue have seen high growth firm increases, with no obvious pattern**. South Yorkshire fits the prevalent pattern.


```{r, fig-highgrowthchange}
#| label: fig-highgrowthchange
#| fig-cap: "Linear trend in high growth firms 2017-2022, expressed in approx. percent change per year"
#| fig-height: 9

firm_slopes <- get_slope_and_se_safely(
  data = bd, 
  region,#slopes will be found within whatever grouping vars are added here
  y = log(count_highgrowth), x = year)
  

#Join geo and slopes for mapping
bd.geo <- la22boundaries %>% 
  rename(region = LAD22NM) %>% 
  left_join(
    bd,
    by = c('region')
    ) %>% 
  left_join(
    firm_slopes,
    by = 'region'
  )

tm_shape(bd.geo %>% mutate(approx_pct_change = slope * 100)) +
  tm_polygons('approx_pct_change', n = 9, palette = "RdBu", border.alpha = 0.4) +
  tm_layout(title = 'High growth firms\nChange over time\nApprox %/yr', legend.outside = T) +
  tm_shape(sy.geo) + 
  tm_borders(lwd = 2, col = 'black') 


```









@fig-turnovermap shows firm 'turnover' (2020-22 average) - a measure of churn that adds births and deaths together, and expresses those as a proportion of total active firms. **Bands of higher turnover/churn** are present in the North, Midlands, across London and in South Wales, with Doncaster standing out in the North and the rest of South Yorkshire not especially low churn.

```{r, fig-turnovermap}
#| label: fig-turnovermap
#| fig-cap: "Turnover: births plus deaths, over count of active firms. 2020-2022 average."
#| fig-height: 9

bd.la <- bd %>% 
  mutate(
  firmefficency = (count_births - count_deaths)/(count_births + count_deaths) * 100,#diffs here quite small so scale to 100
  turnover = (count_births + count_deaths)/(count_active),
  births_over_active_percent = ((count_births)/(count_active))*100,
  deaths_over_active_percent = ((count_deaths)/(count_active))*100,
  highgrowth_over_active_percent = ((count_highgrowth)/(count_active))*100
) %>% 
  group_by(region) %>% 
  mutate(
    #Smoothed variables
    highgrowth_movingav = rollapply(count_highgrowth, 3, mean, align = 'center', fill = NA),
    #derived measures
    firmefficency_movingav = rollapply(firmefficency, 3, mean, align = 'center', fill = NA),
    turnover_movingav = rollapply(turnover, 3, mean, align = 'center', fill = NA),
    births_over_active_percent_movingav = rollapply(births_over_active_percent, 3, mean, align = 'center', fill = NA),
    deaths_over_active_percent_movingav = rollapply(deaths_over_active_percent, 3, mean, align = 'center', fill = NA),
    highgrowth_over_active_percent_movingav = rollapply(highgrowth_over_active_percent, 3, mean, align = 'center', fill = NA)
  ) %>% 
  ungroup()


#Reduce that to get facetted plot of two timepoints for firm efficiency
#Before doing geo-join
bd.la.efficiency.twopoints <- bd.la %>% 
  # select(region,code,year,firmefficency_movingav,births_over_active_percent_movingav,deaths_over_active_percent_movingav,highgrowth_movingav,firmefficency_movingav) %>% 
  filter(year %in% c(2018,2021))#start and end points for 3 year moving av
  


#Change some names for usefulness
bd.la.eff.geo <- la22boundaries %>% 
  rename(region = LAD22NM) %>% 
  left_join(
    bd.la.efficiency.twopoints,
    by = c('region')
  ) %>% 
  mutate(
    year = ifelse(year == 2018, '2017-19 average', '2020-22 average')
  )


#Show only one year here
tm_shape(bd.la.eff.geo %>% filter(year == '2020-22 average')) +
  # tm_polygons('deaths_over_active_percent_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher',
  # tm_polygons('births_over_active_percent_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher',
  tm_polygons('turnover_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher', title = "Firm turnover\n2020-22 average") +
  # tm_polygons('highgrowth_over_active_percent_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher', title = "") +
  # tm_polygons('firmefficency_movingav', n = 9, palette = "RdBu", border.alpha = 0.4, title = "Firm effiency\n0 = equal births + deaths\n 1 = all births no deaths\n-1 = all deaths no births") +
  tm_layout(title = '', legend.outside = T) +
  # tm_facets(by = 'year') +
  tm_shape(sy.geo) + #after facets or doesn't work
  tm_borders(lwd = 2, col = 'black') 
```




@fig-firmefficiencymap shows 'firm efficiency' - the balance of births & deaths - comparing two time points. Values of +100 would be 'all births no deaths', -100 would be 'all deaths no births', and zero would be 'births and deaths balance'. This map shows how in the latter time period (2020-22) there has been a **marked shift to more deaths than births in the South** compared to earlier.

```{r, fig-firmefficiencymap}
#| label: fig-firmefficiencymap
#| fig-cap: "Firm efficiency, comparing 2017-19 av with 2020-22 av."
#| fig-height: 17


tm_shape(bd.la.eff.geo) +
  # tm_polygons('deaths_over_active_percent_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher',
  # tm_polygons('births_over_active_percent_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher',
  # tm_polygons('turnover_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher',
  # tm_polygons('highgrowth_over_active_percent_movingav', n = 9, palette = "Blues", border.alpha = 0.4, style = 'fisher', title = "") +
  tm_polygons('firmefficency_movingav', n = 9, palette = "RdBu", border.alpha = 0.4, title = "Firm effiency\n0 = equal births/deaths\n 100 = all births\n-100 = all deaths") +
  tm_layout(title = '', legend.outside = T) +
  tm_facets(by = 'year', ncol = 1) +
  tm_shape(sy.geo) + #after facets or doesn't work
  tm_borders(lwd = 2, col = 'black') 


```




















```{r eval = F}
#CUTTINZ

#```{r, birthsdeaths_MCA_twoaxis}
##| label: fig-birthsdeaths2axis_MCA_twoaxis
##| fig-cap: "For MCAs: births vs deaths on separate axes as a percent of all active firms, 3 year average 2017/19 (dot) vs 2020/22 (arrow end)."

#Arrows showing births and deaths against each other over time, smoothed


#Arrow start and end points need to be on same row
# bd.mca.wide <- bd.mca %>% 
#   filter(year %in% c(2018,2021)) %>% 
#   select(MCA,year,firmefficency:deaths_over_active_percent_movingav) %>% 
#   pivot_wider(names_from = year, values_from = firmefficency:deaths_over_active_percent_movingav)
# 
# ggplot(
#   bd.mca.wide %>% mutate(SY = MCA == "South Yorkshire"),
#   aes(colour = fct_reorder(MCA,-turnover_movingav_2021))
# ) +
#   geom_segment(aes(size = SY, x = births_over_active_percent_movingav_2018, y = deaths_over_active_percent_movingav_2018 ,
#                    xend = births_over_active_percent_movingav_2021, yend = deaths_over_active_percent_movingav_2021),
#                arrow = arrow(length = unit(0.5, "cm"))) +
#   # geom_segment(aes(x = births_over_active_percent_movingav_2018, y = deaths_over_active_percent_movingav_2018 ,
#   #                  xend = births_over_active_percent_movingav_2021, yend = deaths_over_active_percent_movingav_2021),
#   #              arrow = arrow(length = unit(0.5, "cm")),
#   #              size = 1) +
#   scale_color_brewer(palette = 'Paired') +
#   scale_size_manual(values = c(1,3)) +
#   geom_point(aes(x = births_over_active_percent_movingav_2018, y = deaths_over_active_percent_movingav_2018), size = 3) +
#   geom_abline(slope = 1, intercept = 0) +
#   # scale_y_log10() +
#   # scale_x_log10() +
#   coord_fixed() +
#   labs(colour = 'MCA') +
#   xlab('Births as % of all active firms') +
#   ylab('Deaths as % of all active firms') +
#   guides(size = F)






# ```{r, birthsdeaths_core}
# #| label: fig-birthsdeaths2axis_core
# #| fig-cap: "For core cities plus BDR: births vs deaths on separate axes as a percent of all active firms, 3 year average 2017/19 (dot) vs 2020/22 (arrow end)."

#And for core cities?
# bd.core <- bd %>%
#   # filter(corecity == 'Core city') %>% 
#   filter(corecity == 'Core city' | grepl('barns|doncaster|rotherh',region,ignore.case=T)) %>% #version that includes other places in SY
#   filter(!grepl('cardiff|belfast',region,ignore.case=T)) %>% 
#   mutate(
#     firmefficency = (count_births - count_deaths)/(count_births + count_deaths) * 100,#diffs here quite small so scale to 100
#     turnover = (count_births + count_deaths)/(count_active),
#     births_over_active_percent = ((count_births)/(count_active))*100,
#     deaths_over_active_percent = ((count_deaths)/(count_active))*100
#   ) %>% 
#   group_by(region) %>% 
#   mutate(
#     #Smoothed variables
#     highgrowth_movingav = rollapply(count_highgrowth, 3, mean, align = 'center', fill = NA),
#     #derived measures
#     firmefficency_movingav = rollapply(firmefficency, 3, mean, align = 'center', fill = NA),
#     turnover_movingav = rollapply(turnover, 3, mean, align = 'center', fill = NA),
#     births_over_active_percent_movingav = rollapply(births_over_active_percent, 3, mean, align = 'center', fill = NA),
#     deaths_over_active_percent_movingav = rollapply(deaths_over_active_percent, 3, mean, align = 'center', fill = NA)
#   ) %>% 
#   ungroup()
# 
# #Arrow start and end points need to be on same row
# bd.core.wide <- bd.core %>% 
#   filter(year %in% c(2018,2021)) %>% 
#   select(region,year,firmefficency:deaths_over_active_percent_movingav) %>% 
#   pivot_wider(names_from = year, values_from = firmefficency:deaths_over_active_percent_movingav)
# 
# 
# 
# ggplot(
#   bd.core.wide %>% mutate(SBDR = grepl('barnsley|doncast|rotherh|sheffield',region,ignore.case=T)),
#   aes(colour = fct_reorder(region,-turnover_movingav_2021))
# ) +
#   geom_segment(aes(size = SBDR, x = births_over_active_percent_movingav_2018, y = deaths_over_active_percent_movingav_2018 ,
#                    xend = births_over_active_percent_movingav_2021, yend = deaths_over_active_percent_movingav_2021),
#                arrow = arrow(length = unit(0.5, "cm"))) +
#   # geom_segment(aes(x = births_over_active_percent_movingav_2018, y = deaths_over_active_percent_movingav_2018 ,
#   #                  xend = births_over_active_percent_movingav_2021, yend = deaths_over_active_percent_movingav_2021),
#   #              arrow = arrow(length = unit(0.5, "cm")),
#   #              size = 1) +
#   scale_color_brewer(palette = 'Paired') +
#   scale_size_manual(values = c(1,3)) +
#   geom_point(aes(x = births_over_active_percent_movingav_2018, y = deaths_over_active_percent_movingav_2018), size = 3) +
#   geom_abline(slope = 1, intercept = 0) +
#   # scale_y_log10() +
#   # scale_x_log10() +
#   coord_fixed() +
#   labs(colour = 'MCA') +
#   xlab('Births as % of all active firms') +
#   ylab('Deaths as % of all active firms') +
#   guides(size = F)




```














